---
title: "BudgetBookAnalysis"
author: "Raka Adinugraha"
date: "November 21, 2019"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r load Packages, include = FALSE}
library(tidyverse)
library(lubridate)
library(scales)
library(forcats)
library(TSclust)
library(reshape2)
library(plotly)
```

## About the data
This data is a record of my personal spending from January 2019 to August 2019.
During this period, I record every spending that occurs in every account that I have which is, cash money, bank account, and various saving box.
The data itself was exported from 'My Budget Book' android app in my phone. 
## Purpose
The purpose of this analysis is to see the spending trend over this period and make it as a guidefor future spending decisions to improve finances conditions. Other than that, it also uses as an indicator of how good or not I at doing the records of the spending.

## Processing the data
The datasets turned into a tibble. Preset Date and Amount as date and number type respectively. The rest of the column will become character type by default. 

```{r read data}
budget <- read_csv2("January-august_2019.csv", col_types = cols(
  Date = col_date(format = "%m/%d/%Y"),
  Amount = col_number()
))
head(budget)
```

Upon looking to the dataset, we can see there is some element of it we can tweak. First we can see that 'Main category' column name have a space in it, that in my experience could be problematic when try to use it for plotting, so we need to change it. We also add new 'Month_Year' column that will help our further analysis.
```{r change column name & add Month_Year column}
budget <- rename(budget, Main_category = 'Main category')
budget$Month_Year <- format(as_date(budget$Date), '%Y-%m')
head(budget)
```


## GitHub Documents

This is an R Markdown format used for publishing markdown documents to GitHub. When you click the **Knit** button all R code chunks are run and a markdown file (.md) suitable for publishing to GitHub is generated.

## Including Code

You can include R code in the document as follows:

```{r cars}
summary(cars)
```

## Including Plots

You can also embed plots, for example:

```{r pressure, echo=FALSE}
plot(pressure)
```

Note that the `echo = FALSE` parameter was added to the code chunk to prevent printing of the R code that generated the plot.
